STJ2012 analysis

### 2016-08-10 analysis/STJ2012_analysis.Rmd
both 97% OTU and 97% within-sample OTU analyses

options(scipen = 3, digits = 2)
library(phyloseq); library(vegan); library(multcompView)

taxcolors <- matrix(c(rgb(141, 211, 199, maxColorValue=255), rgb(255, 255, 179, maxColorValue=255), 
                      rgb(190, 186, 218, maxColorValue=255), rgb(251, 128, 114, maxColorValue=255), 
                      rgb(253, 180, 98, maxColorValue=255)), 
                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF")))
```

@@

# Data pre-processing
## Import dataset
The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object to facilitate downstream analyses.
```{r data_import, cache=T}
# Import OTU tables
OTU97 <- otu_table(read.table("data/97_otu_table.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
OTU100 <- otu_table(read.table("data/100_otu_table.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy
TAX97 <- tax_table(as.matrix(read.table("data/97_tax_assignments.txt", sep="\t", row.names=2,
                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
TAX100 <- tax_table(as.matrix(read.table("data/100_tax_assignments.txt", sep="\t", row.names=2,
                                         col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
# Build phyloseq objects
phy97 <- phyloseq(OTU97, TAX97, SAM)
phy100 <- phyloseq(OTU100, TAX100, SAM)
```

#### Descriptive statistics and distribution of OTU counts
```{r phy_stats, fig.height=4, fig.width=8}
# Compute summary statistics
stats <- function(phy) {
  return(
    data.frame(rangeotus=paste0(range(taxa_sums(phy)), collapse=" - "),
               nsingles=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
               rangerps=paste0(range(sample_sums(phy)), collapse=" - "),
               amrps=paste(as.integer(mean(sample_sums(phy))), 
                           as.integer(sd(sample_sums(phy))), sep=" ± "),
               gmrps=paste(as.integer(exp(mean(log(sample_sums(phy))))),
                           as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "))
  )
}
stats97 <- stats(phy97)
stats100 <- stats(phy100)

# Create and plot histograms
phy <- phy97
taxhist97 <- hist(log10(taxa_sums(phy97)), plot=F)
samhist97 <- hist(log10(sample_sums(phy97)), plot=F)

@@

par(mfrow=c(2, 2), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
```

|                                  |    97% OTUs               |    100% OTUs                |
| -------------------------------- | ------------------------- | --------------------------- |
| Total count in OTU table         | `r sum(otu_table(phy97))` |  `r sum(otu_table(phy100))` |
| Number of OTUs                   | `r ntaxa(phy97)`          |  `r ntaxa(phy100)`          |
| Range of OTU counts              | `r stats97$rangeotus`     | `r stats100$rangeotus`      |
| Number of singleton OTUs         | `r stats97$nsingles`      | `r stats100$nsingles`       |
| Number of samples                | `r nsamples(phy97)`       |  `r nsamples(phy100)`       |
| Range of reads per sample        | `r stats97$rangerps`      | `r stats100$rangerps`       |
| Mean (±SD) reads per sample (arithmetic) | `r stats97$amrps` | `r stats100$amrps`          |
| Mean (±SD) reads per sample (geometric)  | `r stats97$gmrps` | `r stats100$gmrps`          |

## Filter dataset
#### Filter samples by minimum count
```{r filter_samples}
# Set threshold number of reads
sn <- 10
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97)>=sn, phy97)
phy100.f <- prune_samples(sample_sums(phy100)>=sn, phy100)

@@

* Number of samples remaining in 97% OTU dataset: `r nsamples(phy97.f) `
* Number of samples remaining in 100% OTU dataset: `r nsamples(phy100.f) `

#### Filter OTUs by minimum count
```{r filter_taxa}
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]  
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)  
phy100.f <- prune_taxa(names(taxa100), phy100.f)  
```

* Number of OTUs remaining in 97% dataset: `r ntaxa(phy97.f)`
* Number of OTUs remaining in 97% dataset: `r ntaxa(phy100.f)`

#### Filtered data: descriptive statistics and distribution of OTUs
```{r phy.f_histograms, fig.height=4, fig.width=8}
# Compute summary statistics
stats97.f <- stats(phy97.f)
stats100.f <- stats(phy100.f)

# Create and plot histograms
taxhist97 <- hist(log10(taxa_sums(phy97.f)), plot=F)
taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
samhist97 <- hist(log10(sample_sums(phy97.f)), plot=F)
samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)

par(mfrow=c(2, 2), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
```

|                                  |    97% OTUs                     |    100% OTUs                |
| -------------------------------- | ------------------------------- | --------------------------- |
| Total count in OTU table         | `r sum(otu_table(phy97.f))`     | `r sum(otu_table(phy100.f))`|
| Number of OTUs                   | `r ntaxa(phy97.f)`              | `r ntaxa(phy100.f)`         |
| Range of OTU counts              | `r stats97.f$rangeotus`         | `r stats100.f$rangeotus`    |
| Number of singleton OTUs         | `r stats97.f$nsingles`          | `r stats100.f$nsingles`     |
| Number of samples                | `r nsamples(phy97.f)`           | `r nsamples(phy100.f)`      |
| Range of reads per sample        | `r stats97.f$rangerps`          | `r stats100.f$rangerps`     |
| Mean (±SD) reads per sample (arithmetic) | `r stats97.f$amrps`     | `r stats100.f$amrps`        |
| Mean (±SD) reads per sample (geometric)  | `r stats97.f$gmrps`     | `r stats100.f$gmrps`        |

## Transform count data
```{r transform}
# Convert to proportion (relative abundance)
phy97.f.p <- transform_sample_counts(phy97.f, function(x) x/sum(x))
phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy97.f.t <- transform_sample_counts(phy97.f, transform)  # Transform data
phy100.f.t <- transform_sample_counts(phy100.f, transform)
```

@@

@@

sppdiffs97 <- sppdiffs(phy97.f.t)
sppdiffs100 <- sppdiffs(phy100.f.t)

sppdiffs <- data.frame("97% OTUs"=sppdiffs97$Letters, "100% OTUs"=sppdiffs100$Letters, check.names=F)
knitr::kable(sppdiffs, caption="Species not sharing a letter are significantly different (p < 0.05)")
```

@@

```{r shorediffs, cache=T}
# Compute within- and between-shore statistics for each Species
shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy97.f.t))$Species), 
                                              c("Species", "n", "overall", "within", 
                                                "between", "R2", "bd", "p")), ncol=8, nrow=10))
for (i in 1:nlevels(data.frame(sample_data(phy97.f.t))$Species)) {
  sp <- levels(data.frame(sample_data(phy97.f.t))$Species)[i]
  shorestats$Species[i] <- sp
  phy97.f.t.species <- subset_samples(phy97.f.t, Species==sp)  # Set temp phyloseq object
  shorestats$n[i] <- nsamples(phy97.f.t.species)
  md <- meandist(phyloseq::distance(phy97.f.t.species, "bray"), 
                 sample_data(phy97.f.t.species)$Shore)
  shorestats$within[i] <- summary(md)$W  # Weighted mean dissimilarity within both Pools
  shorestats$between[i] <- summary(md)$B  # Mean dissimilarity between Pools
  shorestats$overall[i] <- summary(md)$D  # Overall dissimilarity
  if (nlevels(as(sample_data(phy97.f.t.species), "data.frame")$Shore) > 1) {
    set.seed(70235)
    permanova <- adonis(phyloseq::distance(phy97.f.t.species, "bray") ~ Shore, 
                        data=as(sample_data(phy97.f.t.species), "data.frame"), permutations=999)
    shorestats$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
    shorestats$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
    bd <- betadisper(phyloseq::distance(phy97.f.t.species, "bray"),
                     group=as(sample_data(phy97.f.t.species), "data.frame")$Shore)
    shorestats$bd[i] <- TukeyHSD(bd)$group[4]
  } else {
    shorestats$R2[i] <- NA
    shorestats$p[i] <- NA
    shorestats$bd[i] <- NA
  }
}
knitr::kable(shorestats)  # dendrogyra, porites furcata, and siderastrea siderea differ by shore
```

## Beta diversity
Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species group. Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test. This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
```{r betadisper, cache=T, fig.width=4, fig.height=4}
# Calculate distances to species centroids for each sample in principal coordinate space
sambd <- betadisper(d=vegdist(t(otu_table(phy97.f.t)), method="bray"), 
                    group=data.frame(sample_data(phy97.f.t))$Species, type="centroid",
                    bias.adjust=TRUE)
# Calculate each species' mean distance to centroid and standard error
sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
                      FUN=function(x) sd(x)/sqrt(length(x)))$x
# Determine order of decreasing mean dispersion
sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
# Update betadisper object with species in decreasing order of mean dispersion
sambd <- betadisper(d=vegdist(t(otu_table(phy97.f.t)), method="bray"), 
                    group=factor(data.frame(sample_data(phy97.f.t))$Species, levels=as.character(sambdsumm.ord$group)), type="centroid", bias.adjust=TRUE)
# Use permutations to perform pairwise comparisons of group mean dispersions
sampt <- permutest(sambd, pairwise=T)
saml <- multcompLetters(sampt$pairwise$permuted)

# Figure: Distance to centroids
par(mfrow=c(1,1), mar=c(3.5,4,1,1), lwd=1)
with(sambdsumm.ord, {
  plot(mean, type="n", ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
  points(1:10, mean, cex=1, pch=21, bg="white")
  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
       labels=levels(SAM$Species)[order(sambdsumm$mean, decreasing=T)])
  text(1:10, mean + se + 0.03, labels=saml$Letters[as.character(group)], cex=0.5)
})
```

```{r strigosa}
dstr97 <- subset_samples(phy97.f.p, Species=="strigosa")
plot_bar(dstr97, x="Sample", y="Abundance", fill="OTU")
plot_bar(dstr97, x="Sample", y="Abundance", fill="Subtype")
# D strigosa hosted B1, B24, B35, B38, and C3 in Finney et al.... B25, B35, and B38 are missing from ref database.
# Also try 100% OTUs -- must filter more first...
dstr100.f <- subset_samples(phy100.f, Species=="strigosa")
dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
#dstr100.f.p.f <- prune_taxa(rownames(otu_table(dstr100.f.p))[which(rowSums(otu_table(dstr100.f.p)) > 0.01)],
#                            dstr100.f.p)
#  
#plot_bar(dstr100.f.p.f, x="Sample", y="Abundance", fill="OTU")
#plot_bar(dstr100, x="Sample", y="Abundance", fill="Subtype")
#sample_names(dstr100.f)
#subset_samples(dstr100, Sample=="7")
#otu_table(dstr100.f)
#dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
#colSums(otu_table(dstr100.f.p)[rowSums(otu_table(dstr100.f.p)) > 0.1, ])

>>>>

options(scipen = 3, digits = 2)

library(phyloseq); library(vegan); library(multcompView); library(reshape2); library(igraph); library(stringr)

taxcolors <- matrix(c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462"), 
                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG")))
```

@@

# Data pre-processing
## Import dataset
The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object to facilitate downstream analyses.
```{r data_import, cache=T}
# Import OTU tables
OTU97 <- otu_table(read.table("data/97_otus.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
OTU97bs <- otu_table(read.table("data/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
OTU100 <- otu_table(read.table("data/100_otus.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy
TAX97 <- tax_table(as.matrix(read.table("data/97_otus_tax_assignments.txt", sep="\t", row.names=2,
                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
TAX97bs <- tax_table(as.matrix(read.table("data/97_otus_bysample_tax_assignments.txt", sep="\t", row.names=2,
                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
TAX100 <- tax_table(as.matrix(read.table("data/100_otus_tax_assignments.txt", sep="\t", row.names=2,
                                         col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
# Build phyloseq objects
phy97 <- phyloseq(OTU97, TAX97, SAM)
phy97bs <- phyloseq(OTU97bs, TAX97bs, SAM)
phy100 <- phyloseq(OTU100, TAX100, SAM)
```

#### Descriptive statistics and distribution of OTU counts
```{r phy_stats, cache=T, fig.height=4, fig.width=8}
# Compute summary statistics
stats <- function(phy) {
  return(
    data.frame(
      'Total count in OTU table'=sum(otu_table(phy)),
      'Number of OTUs'=ntaxa(phy),
      'Range of OTU counts'=paste0(range(taxa_sums(phy)), collapse=" - "),
      'Number of singleton OTUs'=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
      'Number of samples'=nsamples(phy),
      'Range of reads per sample'=paste0(range(sample_sums(phy)), collapse=" - "),
      'Arithmetic mean (±SD) reads per sample'=paste(as.integer(mean(sample_sums(phy))), 
                                                     as.integer(sd(sample_sums(phy))), sep=" ± "),
      'Geometric mean (±SD) reads per sample'=paste(as.integer(exp(mean(log(sample_sums(phy))))),
                                                    as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "),
      check.names=F)
  )
}
stats97 <- data.frame(`97% OTUs`=t(stats(phy97)), check.names=F)
stats97bs <- data.frame(`97% within-sample OTUs`=t(stats(phy97bs)), check.names=F)
stats100 <- data.frame(`100% OTUs`=t(stats(phy100)), check.names=F)

# Create and plot histograms
taxhist97 <- hist(log10(taxa_sums(phy97)), plot=F)
samhist97 <- hist(log10(sample_sums(phy97)), plot=F)

taxhist97bs <- hist(log10(taxa_sums(phy97bs)), plot=F)
samhist97bs <- hist(log10(sample_sums(phy97bs)), plot=F)

@@

par(mfrow=c(2, 3), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist97bs, col="black", main="97% within-sample OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist97bs, col="black", main="97% within-sample OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)

# Create stats table
knitr::kable(cbind(stats97, stats97bs, stats100))
```

## Filter dataset
#### Filter samples by minimum count
```{r filter_samples, cache=T}
# Set threshold number of reads
sn <- 0
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97)>=sn, phy97)
phy97bs.f <- prune_samples(sample_sums(phy97bs)>=sn, phy97bs)
phy100.f <- prune_samples(sample_sums(phy100)>=sn, phy100)

@@

#### Filter OTUs by minimum count
```{r filter_taxa}
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]  
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
phy100.f <- prune_taxa(names(taxa100), phy100.f)

# Refilter samples in case removal of OTUs caused any samples to drop below threshold
sn <- 200
phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
```

#### Filtered data: descriptive statistics and distribution of OTUs
```{r phy.f_histograms, fig.height=4, fig.width=8}
# Compute summary statistics
stats97.f <- data.frame(`97% OTUs`=t(stats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% within-sample OTUs`=t(stats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(stats(phy100.f)), check.names=F)

# Create and plot histograms
taxhist97 <- hist(log10(taxa_sums(phy97.f)), plot=F)
taxhist97bs <- hist(log10(taxa_sums(phy97bs.f)), plot=F)
taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
samhist97 <- hist(log10(sample_sums(phy97.f)), plot=F)
samhist97bs <- hist(log10(sample_sums(phy97bs.f)), plot=F)
samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)

par(mfrow=c(2, 3), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist97bs, col="black", main="97% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist97bs, col="black", main="97% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)

# Create stats table
knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f))
```

## Transform count data
```{r transform}
# Convert to proportion (relative abundance)
phy97.f.p <- transform_sample_counts(phy97.f, function(x) x/sum(x))
phy97bs.f.p <- transform_sample_counts(phy97bs.f, function(x) x/sum(x))
phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy97.f.t <- transform_sample_counts(phy97.f, transform)  # Transform data
phy97bs.f.t <- transform_sample_counts(phy97bs.f, transform) 
phy100.f.t <- transform_sample_counts(phy100.f, transform)
```

@@

# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade], legend=T)

# Plot composition of 97% within-sample OTUs colored by OTU
composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)

@@

sppdiffs97 <- sppdiffs(phy97.f.t)
sppdiffs97bs <- sppdiffs(phy97bs.f.t)
sppdiffs100 <- sppdiffs(phy100.f.t)

sppdiffs <- data.frame("97% OTUs"=sppdiffs97$Letters, "97% within-sample OTUs"=sppdiffs97bs$Letters,
                       "100% OTUs"=sppdiffs100$Letters, check.names=F)
knitr::kable(sppdiffs, caption="Species not sharing a letter are significantly different (p < 0.05)")
```

@@

```{r shorediffs, cache=T}
shorediffs <- function(phy) {
  # Compute within- and between-shore statistics for each Species
  shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy))$Species), 
                                                c("Species", "n", "overall", "within", 
                                                  "between", "R2", "bd", "p")), ncol=8, nrow=10))
  for (i in 1:nlevels(data.frame(sample_data(phy))$Species)) {
    sp <<- levels(data.frame(sample_data(phy))$Species)[i]
    shorestats$Species[i] <- sp
    phy.species <- subset_samples(phy, Species==sp)  # Set temp phyloseq object
    shorestats$n[i] <- nsamples(phy.species)
    md <- meandist(phyloseq::distance(phy.species, "bray"), 
                   sample_data(phy.species)$Shore)
    shorestats$within[i] <- summary(md)$W  # Weighted mean dissimilarity within both Pools
    shorestats$between[i] <- summary(md)$B  # Mean dissimilarity between Pools
    shorestats$overall[i] <- summary(md)$D  # Overall dissimilarity
    if (nlevels(as(sample_data(phy.species), "data.frame")$Shore) > 1) {
      #set.seed(70235)
      set.seed(43789)
      permanova <- adonis(phyloseq::distance(phy.species, "bray") ~ Shore, 
                          data=as(sample_data(phy.species), "data.frame"), permutations=999)
      shorestats$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
      shorestats$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
      bd <- betadisper(phyloseq::distance(phy.species, "bray"),
                       group=as(sample_data(phy.species), "data.frame")$Shore)
      shorestats$bd[i] <- TukeyHSD(bd)$group[4]
    } else {
      shorestats$R2[i] <- NA
      shorestats$p[i] <- NA
      shorestats$bd[i] <- NA
    }
  }
  return(shorestats)
}

shorestats97 <- shorediffs(phy97.f.t)
shorestats97bs <- shorediffs(phy97bs.f.t)
shorediffs(phy97bs.f.p)
shorestats100 <- shorediffs(phy100.f.t)

knitr::kable(shorestats97, row.names=F, 
             caption="97% OTU data: differences within and between shore for each species")
knitr::kable(shorestats97bs, row.names=F, 
             caption="97% within-sample OTU data: differences within and between shore for each species")
knitr::kable(shorestats100, row.names=F, 
             caption="100% OTU data: differences within and between shore for each species")
```

## Beta diversity
Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species group. Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test. This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
```{r betadisper, cache=T, fig.width=10, fig.height=4}
betad <- function(phy) {
  # Calculate distances to species centroids for each sample in principal coordinate space
  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
                      group=data.frame(sample_data(phy))$Species, type="centroid",
                      bias.adjust=TRUE)
  # Calculate each species' mean distance to centroid and standard error
  sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
  sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
                            FUN=function(x) sd(x)/sqrt(length(x)))$x
  # Determine order of decreasing mean dispersion
  sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
  # Update betadisper object with species in decreasing order of mean dispersion
  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
                      group=factor(data.frame(sample_data(phy))$Species,
                                   levels=as.character(sambdsumm.ord$group)), 
                      type="centroid", bias.adjust=TRUE)
  # Use permutations to perform pairwise comparisons of group mean dispersions
  sampt <- permutest(sambd, pairwise=T)
  saml <- multcompLetters(sampt$pairwise$permuted)
  # Return results as a list
  return(list(sambdsumm.ord=sambdsumm.ord, sambdsumm=sambdsumm, saml=saml))
}

betad97 <- betad(phy97.f.t)
betad97bs <- betad(phy97bs.f.t)
betad100 <- betad(phy100.f.t)

# Figure: Distance to centroids
par(mfrow=c(1,3), mar=c(3.5,4,2,1), lwd=1)
with(betad97$sambdsumm.ord, {
  plot(mean, type="n", main="97% OTUs",
       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
  points(1:10, mean, cex=1, pch=21, bg="white")
  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
       labels=levels(SAM$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
  text(1:10, mean + se + 0.03, labels=betad97$saml$Letters[as.character(group)], cex=0.5)
})
with(betad97bs$sambdsumm.ord, {
  plot(mean, type="n", main="97% within-sample OTUs",
       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
  points(1:10, mean, cex=1, pch=21, bg="white")
  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
       labels=levels(SAM$Species)[order(betad97bs$sambdsumm$mean, decreasing=T)])
  text(1:10, mean + se + 0.03, labels=betad97bs$saml$Letters[as.character(group)], cex=0.5)
})
with(betad100$sambdsumm.ord, {
  plot(mean, type="n", main="100% OTUs",
       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
  points(1:10, mean, cex=1, pch=21, bg="white")
  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
       labels=levels(SAM$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
  text(1:10, mean + se + 0.03, labels=betad100$saml$Letters[as.character(group)], cex=0.5)
})
```

```{r otu_barplots_custom, fig.width=8, fig.height=8}
otubarplot <- function(phy, main) {
  samdat <- data.frame(sample_data(phy))
  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
                                                 rownames(samdat)])
  typerelabund <- typerelabund[, rownames(sample_data(phy)[with(sample_data(phy), 
                                                                order(Shore, Site, InputFileName))])]
  # Get info for plotting OTU names and blast hits on top of barplot
  blasthits <- as.character(data.frame(tax_table(phy))[order(data.frame(tax_table(phy))$Subtype), "Subtype"])
  names <- as.character(rownames(data.frame(tax_table(phy)))[order(data.frame(tax_table(phy))$Subtype)])
  samples <- colnames(typerelabund)
  shores <- data.frame(sample_data(phy))[samples, "Shore"]
  sites <- data.frame(sample_data(phy))[samples, "Site"]
  divides <- apply(typerelabund, 2, cumsum)
  heights <- diff(divides)
  heights[heights < 0.04] <- NA
  # Plot barplot and OTU names and blast hits
  bars <- barplot(typerelabund, col=rainbow(ntaxa(phy)), las=1, space=0.1, xaxs="i",
                  xlab="", ylab="", xaxt="n")
  text(bars, -0.06, labels=samples, xpd=T)
  text(bars, -0.10, labels=shores, xpd=T)
  text(bars, -0.14, labels=substr(sites, 1, 2), xpd=T)
  text(bars[1]/4, c(-0.06, -0.10, -0.14), labels=c("sample:", "shore:", "site:"), xpd=T, pos=2)
  mtext(side=2, text = "Relative Abundance", line=2)
  for (i in 1:length(bars)) {
    text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
         divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
         labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
         cex=0.75)
  }
  mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
}
```
